Anony\emph{mice}d shareable data: Using \emph{mice} to create and analyze multiply imputed synthetic data sets

### 2021-10-06 manuscript/Manuscript.tex

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Synthetic data sets can greatly improve the dissemination of
data and further analysis of private data. Generating and analyzing
synthetic data sets is straightforward, yet adopting a synthetic data
analysis pipeline is often ignored by applied researchers. We outline a
simple procedure for generating and alayzing synthetic data sets with
\texttt{mice} in \texttt{R}. We demonstrate in simulation that the
analysis results obtained on synthetic data obtained with \texttt{mice}
yields unbiased and valid inferences, and leads to synthetic records
that cannot be distinguished from the true data records. The ease of use
when synthesizing data with \texttt{mice}, together with the validity of
inferences obtained through this procedure opens up a wealth of
possibilities for data dissemination and further research of initially
private data.}

@@

Open science, including open data, has been marked as the future of
science \citep{gewin_data_2016}, and the advantages of publicly
available research data are numerous
\citep{molloy_open_2011, walport_brest_sharing_2011}. Collecting
research data requires an enormous investment both in terms of time and
monetary resources. Openly accessible research data bears the potential
of increasing the scientific returns for the same data collection
effort. Additionally, the fact that public funds are used for data
collection results in increasing demand for the collected data.
Nevertheless, the possibilities to distribute research data directly are
often very limited due to restrictions on data privacy and data
confidentiality. Although these regulations are much needed, privacy
constraints are also ranked among the toughest challenges to overcome in
the advancement of modern day social science research
\citep{lazer_life_2009}.

@@

Potentially, this approach could fulfill the needs for openly accessibly
data, without running into barriers with regard to privacy and
confidentiality constraints. However, there is no such thing as a free
lunch: data collectors have to put effort in creating high-quality
synthetic data. Also, the quality of the synthetic data is highly
dependent on the imputation models, and using flawed models to generate
synthetic data might bias subsequent analyses. Conversely, if the models
used to create the synthetic data are able to preserve the relationships
between the variables as in the original data, the synthetic data can be
nearly as informative as the observed data. Thus, to fully exploit the
benefits of synthetic data, the effort to actually create these
high-quality data sets should be kept at a minimum.

To mitigate additional effort of creating synthetic data sets on behalf
of the researcher, software aimed at multiple imputation of missing data
can be employed. Especially if researchers used this software at an
earlier stage in the research process, or acquired familiarity with it
during earlier projects, the additional burden of creating synthetic
data sets is relatively small. The R-package \texttt{mice} \citep{mice}
implements multiple imputation of missing data in a straightforward and
user-friendly manner. However, the functionality of \texttt{mice} is not
restricted to the imputation of missing data, but allows to impute any
value in the data: even observed values. Consequently, \texttt{mice} can
be utilized for the creation of multiply imputed synthetic data sets.

After creating the multiply imputed synthetic data sets, the goal is to
obtain valid statistical inferences in the spirit of
\citet{rubin_multiple_1987} and \citet{neyman1934}. In the missing data
framework, this is done by performing statistical analyses on all
imputed data sets, and pooling the results of the analyses according to
Rubin's rules \citep[pp.76]{rubin_multiple_1987}. In the synthetic data
framework, the same procedure is followed, but with a slight twist:
there are no values that remain constant over the synthetic data sets.
The procedure of drawing valid inferences from multiple synthetic data
sets is therefore slightly different.

@@

The \texttt{mice} package \citep{mice} in \texttt{R} \citep{Rproject}
has been developed for multiple imputation of missing data. In this
context, the aim is to replace missing values due to nonresponse by
plausible values from the posterior predictive distribution of the
variable containing the missings. Doing so, \texttt{mice} makes use of
fully conditional specification {[}FCS; \citet{vanbuuren_fully_2006}{]},
which breaks down the multivariate distribution of the data
\(\textbf{Y} = (\textbf{Y}_{obs}, \textbf{Y}_{mis})\) into
\(j = 1, 2, \dots, k\) univariate conditional densities, where \(k\)
denotes the number of columns in the data. Using FCS, a model is
constructed for every incomplete variable and the missing values
\(Y_{j, mis}\) are then imputed with draws from the posterior predictive
distribution of \(P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)\) on a
variable-by-variable basis. Note that the predictor matrix \(Y_{-j}\)
may contain yet imputed values from an earlier imputation step, and thus
will be updated after every iteration. This procedure is applied \(m\)
times, resulting in \(m\) completed data sets
\(\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})\),
with \(\textbf{D}^{(l)} = (\textbf{Y}_{obs}, Y^{(l)}_{mis})\).

@@

It is straightforward to extended the imputation approach to generate
synthetic values. Rather than imputing missing data, the observed values
are then replaced by synthetic draws from the posterior predictive
distribution. For simplicity, assume that the data is completely
observed (i.e., \(\textbf{Y} = \textbf{Y}_{obs}\)). Following the
notation of \citet{reiter_raghunathan_multiple_2007}, let for \(n\)
units denote \(Z_i = 1\) if any of the values of unit
\(i = 1, 2, \dots, n\) are to be replaced by imputations, and
\(Z_i = 0\) otherwise, with \(\textbf{Z} = (Z_1, Z_2, \dots, Z_n)\).
Accordingly, the data consists of values that are to be replaced and
values that are to be kept (i.e.,
\(\textbf{Y} = (\textbf{Y}_{rep}, \textbf{Y}_{nrep})\). Now, instead of
imputing \(\textbf{Y}_{mis}\) with draws from the posterior predictive
distribution of \(P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)\) as in the
missing data case, \(\textbf{Y}_{rep}\) is imputed from the posterior
distribution of
\(P(Y^{(l)}_{j, rep} | \textbf{Y}^{(l)}_{-j}, \textbf{Z}, \theta)\),
where \(l\) is an indicator for the synthetic data set
(\(l = 1, 2, \dots, m\)). Note that synthetic values that are imputed at
an earlier step can be used for imputing variable \(j\). This process
results in the synthetic data
\(\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})\).

@@

where the argument \texttt{where} requires a matrix of the same
dimensions as the data, (i.e., a \(n \times k\) matrix) containing
logicals \(z_{ij}\) that indicate which cells are selected to have their
values replaced by draws from the posterior predictive distribution.
This approach allows to \emph{overimpute} a subset of the observed data,
or - as in the above example - the observed data as a whole, resulting
in a data set that partly or completely consists of synthetic data
values.

Choosing an adequate imputation model to impute the data is paramount,
as a flawed imputation model may drastically impact the validity of
inferences. Imputation models should be as flexible as possible to
capture most of the patterns in the data, and to model possibly
unanticipated data characteristics
\citep{murray_multiple_2018, rubin_18years_1996}. Parametric methods,
albeit easy to implement in practice, may be too restrictive to capture
generally complex patterns in the data, especially in the case of
nonlinear relations and interactions between multiple variables.
Classification and regression trees {[}CART;
\citet{breiman_cart_1984}{]} allow to model more complex patterns in the
data, and have therefore been suggested as an appropriate imputation
method
\citep{reiter_cart_2005, burgette_reiter_cart_2010, doove_buuren_recursive_2014}.
Loosely speaking, CART sequentially splits the predictor space into
non-overlapping regions in such a way that the within-region variance is
as small as possible after every split. As such, CART does not impose
any parametric distribution on the data, making it a widely applicable
method that allows for a large variety of relationships within the data
\citep{islr_2013}. Given these appealing characteristics and the call
for the use of flexible methods when multiply imputing data, we will
focus our illustrations and evaluations of \texttt{mice} to method
\texttt{mice.impute.cart()}, realized by:

@@

We demonstrate the suitability of using \texttt{mice} for synthesization
using a simulation study on the \texttt{mice::boys} data set. This data
set consists of the values of \(748\) Dutch boys on the following \(9\)
variables:

@@

Unfortunately, this data set does not differ from the vast majority of
collected data sets, in the sense that it suffers from missing data. For
simplicity, we complete the missing values using the default
\texttt{mice} imputation model for all predictors except \texttt{bmi},
which is passively imputed using its deterministic relation with
\texttt{wgt} and \texttt{hgt}.

@@

To induce sampling variance, 1000 bootstrap samples of the \texttt{boys}
data have been synthesized with \(m = 5\) imputations for every data
cell. Synthetic values are generated using the \texttt{CART} imputation
method for all columns, except for \texttt{bmi}. The deterministic
relation \texttt{bmi} which will be synthesized passively based on the
synthetic values for \texttt{hgt} and \texttt{wgt} to preserve the
relation in the synthetic data. Additional parameters that come with the
use of \texttt{mice.impute.cart()} are the complexity parameter
\texttt{cp} and the minimum number of observations in any terminal node
\texttt{minbucket}, that both constrain the flexibility of the
imputation model. The values of the parameters \texttt{cp} and
\texttt{minbucket} ought to adhere to the call for imputation models
that are as flexible as possible. Appropriate values for these
parameters, as well as the input for the \texttt{predictorMatrix},
depend on the data at hand. In the current example, the complexity
parameter is specified at \texttt{cp\ =\ 1e-08} rather than the default
value \texttt{1e-04}, and the minimum number of observations in each
terminal node is set at \texttt{minbucket} \(= 3\) rather than the
default value \(5\). By allowing for more complexity in the imputation
model, bias in the estimates from the synthetic data set is reduced.
Additionally, since the synthesis pattern is monotone, the number of
iterations can be set to \texttt{maxit\ =\ 1}.

To assess the performance of \texttt{mice} for synthesizing data, we
compare the bootstrapped samples with the synthetic versions of these
bootstrapped samples. Specifically, univariate descriptive statistics,
the correlation matrix, and two linear regression models as well as one
ordered logistic regression model will be considered. Subsequently, the
bias in the parameters and the \(95\%\) confidence interval coverage of
the synthetic data will be examined. Similarly to multiple imputation of
missing data, correct inferences from synthetic data requires correct
pooling over the multiply imputed data sets.

@@

Similarly to the missing data case, variances, and subsequently
confidence intervals, should incorporate the increase in variance that
is due to imputation
\citep{reiter_partially_inference_2003, drechsler_synthetic_2011}. Yet,
the increase in variance due to imputation differs according to whether
missing values are imputed or observed data is overimputed with missing
values. Whereas the variance estimate after imputation of missing data
needs to account for the fact that a certain amount of information in
the data is missing, variance estimation from synthetic data does not
suffer from this issue. The adjusted variance estimate that follows from
using multiple synthetic data sets only suffers from the fact that a
finite number of \(m\) synthetic data sets are used to resemble the
observed data. Hence, the according variance estimate for synthetic data
as developed by \citet{reiter_partially_inference_2003} yields \[
T = \bar{u}_m + \frac{b_m}{m},
\]

@@

\begin{table}[ht]
\caption{Univariate descriptives for the true data and $m=5$ pooled univariate descriptives for the synthetic data over 1000 simulations.}
\centering
\begin{tabular}{rrrrrrrrr}
  \hline
 & n & mean & sd & median & min & max & skew & kurtosis \\ 
  \hline
original age & 748 & 9.16 & 6.89 & 10.50 & 0.04 & 21.18 & -0.03 & -1.56 \\ 
  synthetic age & 748 & 9.15 & 6.89 & 10.49 & 0.04 & 20.96 & -0.03 & -1.55 \\ 
  original hgt & 748 & 131.10 & 46.52 & 145.75 & 50.00 & 198.00 & -0.30 & -1.47 \\ 
  synthetic hgt & 748 & 131.06 & 46.50 & 145.32 & 50.69 & 197.16 & -0.30 & -1.47 \\ 
  original wgt & 748 & 37.12 & 26.03 & 34.55 & 3.14 & 117.40 & 0.38 & -1.03 \\ 
  synthetic wgt & 748 & 37.09 & 26.00 & 34.44 & 3.35 & 112.26 & 0.38 & -1.03 \\ 
  original bmi & 748 & 18.04 & 3.04 & 17.45 & 11.73 & 31.74 & 1.14 & 1.79 \\ 
  synthetic bmi & 748 & 18.05 & 3.08 & 17.48 & 11.49 & 32.37 & 1.11 & 1.85 \\ 
  original hc & 748 & 51.62 & 5.86 & 53.10 & 33.70 & 65.00 & -0.91 & 0.12 \\ 
  synthetic hc & 748 & 51.61 & 5.86 & 53.18 & 34.38 & 62.85 & -0.91 & 0.12 \\ 
  original gen* & 748 & 2.53 & 1.59 & 2.00 & 1.00 & 5.00 & 0.52 & -1.36 \\ 
  synthetic gen* & 748 & 2.53 & 1.59 & 2.00 & 1.00 & 5.00 & 0.52 & -1.35 \\ 
  original phb* & 748 & 2.75 & 1.86 & 2.00 & 1.00 & 6.00 & 0.56 & -1.25 \\ 
  synthetic phb* & 748 & 2.75 & 1.86 & 2.00 & 1.00 & 6.00 & 0.56 & -1.24 \\ 
  original tv & 748 & 8.43 & 8.12 & 3.00 & 1.00 & 25.00 & 0.85 & -0.78 \\ 
  synthetic tv & 748 & 8.42 & 8.11 & 3.19 & 1.00 & 25.00 & 0.85 & -0.77 \\ 
  original reg* & 748 & 3.02 & 1.14 & 3.00 & 1.00 & 5.00 & -0.08 & -0.77 \\ 
  synthetic reg* & 748 & 3.02 & 1.14 & 3.00 & 1.00 & 5.00 & -0.08 & -0.76 \\ 
   \hline
\end{tabular}
\end{table}

We see from Table 1 that the synthetic data estimates closely resemble
the true data estimates.

@@

We see that the correlations obtained from synthetic data are unbiased
with respect to the true data set.

@@

\begin{table}[ht]
\caption{Simulation results for a linear regression model with continuous and ordered categorical predictors. The model evaluated is `hgt ~ age + phb`. Depicted are the true data estimate and the bias from the true data estimate and the coverage rate of the 95\% confidence interval for the bootstrap and synthetic data sets.}
\centering
\begin{tabular}{rrrrrr}
  \hline
  & & \multicolumn{2}{c}{Bootstrap} & \multicolumn{2}{c}{Synthetic}\\
 term & estimate & bias & cov & bias & cov \\ 
  \hline
(Intercept) & 63.087 & -0.001 & 0.970 & 0.405 & 0.958 \\ 
  age & 7.174 & 0.000 & 0.958 & -0.033 & 0.947 \\ 
  phb.L & -12.250 & 0.008 & 0.950 & 0.582 & 0.927 \\ 
  phb.Q & -1.376 & -0.022 & 0.926 & 0.112 & 0.934 \\ 
  phb.C & -3.564 & 0.051 & 0.915 & 0.301 & 0.912 \\ 
  phb\verb|^|4 & -0.431 & 0.016 & 0.930 & 0.106 & 0.940 \\ 
  phb\verb|^|5 & 2.064 & 0.060 & 0.941 & 0.077 & 0.943 \\ 
   \hline
\end{tabular}
\end{table}

We see that the finite nature of the true data set together with the
design-based simulation setup yields slight undercoverage for the
polynomial functions. This finding is observed in both the bootstrap
coverages (i.e.~the fraction of 95\% confidence intervals that cover the
true data parameters) and the synthetic data coverages, indicating that
the undercoverage stems from the simulation setup.

@@

\begin{table}[H]
\caption{Simulation results for a proportional odds logistic regression model with continuous and ordered categorical predictors. The model evaluated is `gen ~ age + hc + reg`. Depicted are the true data estimate and the bias from the true data estimate and the coverage rate of the 95\% confidence interval for the bootstrap and synthetic data sets.}
\centering
\begin{tabular}{rrrrrr}
  \hline
 & & \multicolumn{2}{c}{Bootstrap} & \multicolumn{2}{c}{Synthetic}\\
 term & estimate & bias & cov & bias & cov \\ 
  \hline
  age & 0.461 & 0.004 & 0.942 & 0.002 & 0.939 \\
  hc & -0.188 & -0.000 & 0.929 & -0.004 & 0.945 \\
  regeast & -0.339 & 0.012 & 0.960 & 0.092 & 0.957\\
  regwest & 0.486 & 0.009 & 0.952 & -0.122 & 0.944\\
  regsouth & 0.646 & 0.012 & 0.966 & -0.152 & 0.943 \\
  regcity & -0.069 & 0.012 & 0.940 & 0.001 & 0.972 \\
  G1$|$G2 & -6.322 & 0.032 & 0.934 & -0.254 & 0.946 \\
  G2$|$G3 & -4.501 & 0.052 & 0.936 & -0.246 & 0.945 \\
  G3$|$G4 & -3.842 & 0.058 & 0.937 & -0.244 & 0.948 \\
  G4$|$G5 & -2.639 & 0.064 & 0.936 & -0.253 & 0.947 \\
   \hline
\end{tabular}
\end{table}

These results demonstrate that the synthetic data analysis yields valid
inferences with low bias.

@@

We demonstrate that generating synthetic data sets with \texttt{mice} in
\texttt{R} is a straightforward process that fits well in a data
analysis pipeline. The multiple synthetic sets yield valid inferences
about the true underlying data generating mechanism, thereby capturing
the nature of the original data. This makes the multiple synthetisation
procedure with \texttt{mice} suitable for further dissemination of
synthetic data sets.

>>>>

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Synthetic data sets can greatly improve the dissemination of
data and further analysis of private data. Generating and analyzing
synthetic data sets is straightforward, yet a synthetic data analysis
pipeline is seldom adopted by applied researchers. We outline a simple
procedure for generating and analyzing synthetic data sets with
\texttt{mice} in \texttt{R}. We demonstrate using simulations that the
analysis results obtained on synthetic data yields unbiased and valid
inferences, and leads to synthetic records that cannot be distinguished
from the true data records. The ease of use when synthesizing data with
\texttt{mice}, together with the validity of inferences obtained through
this procedure opens up a wealth of possibilities for data dissemination
and further research of initially private data.}

@@

Open science, including open data, has been marked as the future of
science \citep{gewin_data_2016}, and the advantages of publicly
available research data are numerous
\citep{molloy_open_2011, walport_brest_sharing_2011}. Collecting
research data requires an enormous investment both in terms of time and
monetary resources. Openly accessible research data bears the potential
of increasing the scientific returns for the same data collection
effort. Additionally, the fact that public funds are used for data
collection results in increasing demand for access to the collected
data. Nevertheless, the possibilities to distribute research data
directly are often very limited due to restrictions on data privacy and
data confidentiality. Although these regulations are much needed,
privacy constraints are also ranked among the toughest challenges to
overcome in the advancement of modern day social science research
\citep{lazer_life_2009}.

@@

Potentially, this approach could fulfill the needs for openly accessibly
data, without running into barriers with regard to privacy and
confidentiality constraints. However, there is no such thing as a free
lunch: data collectors have to put effort in creating high-quality
synthetic data. The quality of the synthetic data is highly dependent on
the imputation models, and using flawed models to generate synthetic
data might bias subsequent analyses
\citep{reiter2004simultaneous, grund2021using, jiang2021balancing}.
Conversely, if the models used to create the synthetic data are able to
preserve the relationships between the variables as in the original
data, the synthetic data can be nearly as informative as the observed
data. Thus, to fully exploit the benefits of synthetic data, additional
complications to actually create these high-quality data sets should be
kept at a minimum.

To mitigate unnecessary challenges related to creating synthetic data
sets on behalf of the researcher, software aimed at multiple imputation
of missing data can be employed. Especially if researchers acquired
familiarity with this software during earlier projects, or used it
earlier during the research process, the additional burden of creating
synthetic data sets is relatively small. The R-package \texttt{mice}
\citep{mice} implements multiple imputation of missing data in a
straightforward and user-friendly manner. However, the functionality of
\texttt{mice} is not restricted to the imputation of missing data, but
allows to impute any value in the data: even observed values.
Consequently, \texttt{mice} can be utilized for the creation of multiply
imputed synthetic data sets.

After creating multiply imputed synthetic data sets, the goal is to
obtain valid statistical inferences in the spirit of
\citet{rubin_multiple_1987} and \citet{neyman1934}. In the missing data
framework, this is done by performing statistical analyses on all
imputed data sets, and pooling the results of the analyses according to
Rubin's rules \citep[p.~76]{rubin_multiple_1987}. In the synthetic data
framework, the same procedure is followed, but with a slight twist:
there are no values that remain constant over the synthetic data sets.
The procedure of drawing valid inferences from multiple synthetic data
sets is therefore slightly different.

@@

The \texttt{mice} package \citep{mice} in \texttt{R} \citep{Rproject}
has been developed for multiple imputation to overcome problems related
to nonresponse. In that context, the aim is to replace missing values
with plausible values from the posterior predictive distribution of that
variable. In \texttt{mice}, this is accomplished using fully conditional
specification (FCS) \citep{vanbuuren_fully_2006}, which breaks down the
multivariate distribution of the data
\(\textbf{Y} = (\textbf{Y}_{obs}, \textbf{Y}_{mis})\) into
\(j = 1, 2, \dots, k\) univariate conditional densities, where \(k\)
denotes the number of columns in the data. Using FCS, a model is
constructed for every incomplete variable and the missing values
\(Y_{j, mis}\) are then imputed with draws from the posterior predictive
distribution of \(P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)\) on a
variable-by-variable basis. Note that the predictor matrix \(Y_{-j}\)
may contain yet imputed values from an earlier imputation step, and thus
will be updated after every iteration. This procedure is applied \(m\)
times, resulting in \(m\) completed data sets
\(\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})\),
with \(\textbf{D}^{(l)} = (\textbf{Y}_{obs}, Y^{(l)}_{mis})\).

@@

It is straightforward to extend the imputation approach to generate
synthetic values. Rather than imputing missing data, the observed values
are then replaced by synthetic draws from the posterior predictive
distribution. For simplicity, assume that the data is completely
observed (i.e., \(\textbf{Y} = \textbf{Y}_{obs}\)). Following the
notation of \citet{reiter_raghunathan_multiple_2007}, let for \(n\)
units denote \(Z_i = 1\) if any of the values of unit
\(i = 1, 2, \dots, n\) are to be replaced by imputations, and
\(Z_i = 0\) otherwise, with \(\textbf{Z} = (Z_1, Z_2, \dots, Z_n)\).
Accordingly, the data consists of values that are to be replaced and
values that are to be kept (i.e.,
\(\textbf{Y} = (\textbf{Y}_{rep}, \textbf{Y}_{nrep})\). Now, instead of
imputing \(\textbf{Y}_{mis}\) with draws from the posterior predictive
distribution of \(P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)\) as in the
missing data case, \(\textbf{Y}_{rep}\) is imputed from the posterior
distribution of
\(P(Y^{(l)}_{j, rep} | \textbf{Y}^{(l)}_{-j}, \textbf{Z}, \theta)\),
where \(l\) is an indicator for the synthetic data set
(\(l = 1, 2, \dots, m\)). Note that synthetic values that are imputed at
an earlier step can be used for imputing variable \(j\). This process
results in the synthetic data
\(\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})\).

@@

where the argument \texttt{where} requires a matrix of the same
dimensions as the data, (i.e., a \(n \times k\) matrix) containing
logicals \(z_{ij}\) that indicate which cells are selected to have their
values replaced by draws from the posterior predictive distribution.
This approach allows to overimpute a subset of the observed data, or as
in the above example, the observed data as a whole, resulting in a data
set that partially or completely consists of synthetic data values.

Choosing an adequate imputation model to impute the data is paramount,
as a flawed imputation model may drastically impact the validity of
inferences \citep{grund2021using, jiang2021balancing}. Imputation models
should be as flexible as possible to capture most of the patterns in the
data, and to model possibly unanticipated data characteristics
\citep{murray_multiple_2018, rubin_18years_1996}. Parametric methods,
albeit easy to implement in practice, may be too restrictive to capture
generally complex patterns in the data, especially in the case of
nonlinear relations and interactions between multiple variables.
Classification and regression trees \citep[CART;][]{breiman_cart_1984}
allow to model more complex patterns in the data, and have therefore
been suggested as an appropriate imputation method
\citep{reiter_cart_2005, burgette_reiter_cart_2010, doove_buuren_recursive_2014}.
Loosely speaking, CART sequentially splits the predictor space into
non-overlapping regions in such a way that the within-region variance is
as small as possible after every split. As such, CART does not impose
any parametric distribution on the data, making it a widely applicable
method that allows for a large variety of relationships within the data
\citep{islr_2013}. Given these appealing characteristics and the call
for the use of flexible methods when multiply imputing data, we will
focus our illustrations and evaluations of \texttt{mice} to method
\texttt{mice.impute.cart()}, realized by:

@@

We demonstrate the suitability of using \texttt{mice} for synthetisation
using a simulation study on the \texttt{mice::boys} data set. This data
set consists of the values of \(748\) Dutch boys on the following \(9\)
variables:

@@

Unfortunately, this data set does not differ from the vast majority of
collected data sets, in the sense that it suffers from missing data. For
simplicity, we complete the missing values using the default
\texttt{mice} imputation model for all predictors except \texttt{bmi},
which is passively imputed using its deterministic relation with
\texttt{wgt} and \texttt{hgt}. Specifically, the imputed values are used
to calculate the exact \texttt{bmi} values that correspond with
\texttt{hgt} and \texttt{wgt}.

@@

Usually, one would draw samples from a population that can be
synthesized to evaluate the performance of the synthesization methods.
As we only have access to a single sample, 1000 bootstrap samples of the
\texttt{boys} data have been synthesized with \(m = 5\) imputations for
every data cell to induce an appropriate amount of sampling variance.
Synthetic values are generated using the \texttt{CART} imputation method
for all columns, except for \texttt{bmi}. The deterministic relation
\texttt{bmi} which will be synthesized passively based on the synthetic
values for \texttt{hgt} and \texttt{wgt} to preserve the relation in the
synthetic data. Additional parameters that come with the use of
\texttt{mice.impute.cart()} are the complexity parameter \texttt{cp} and
the minimum number of observations in any terminal node
\texttt{minbucket}, that both constrain the flexibility of the
imputation model. The values of the parameters \texttt{cp} and
\texttt{minbucket} ought to adhere to the call for imputation models
that are as flexible as possible. Appropriate values for these
parameters, as well as the input for the \texttt{predictorMatrix},
depend on the data at hand. In the current example, the complexity
parameter is specified at \texttt{cp\ =\ 1e-08} rather than the default
value \texttt{1e-04}, and the minimum number of observations in each
terminal node is set at \texttt{minbucket} \(= 3\) rather than the
default value \(5\). By allowing for more complexity in the imputation
model, bias in the estimates from the synthetic data set is reduced.
Additionally, since the synthesis pattern is monotone, the number of
iterations can be set to \texttt{maxit\ =\ 1} \citep[e.g.,][Ch.
3]{drechsler_synthetic_2011}.

To assess the performance of \texttt{mice} for synthesizing data, we
compare the bootstrapped samples with the synthetic versions of these
bootstrapped samples. Specifically, univariate descriptive statistics,
the correlation matrix, and two linear regression models as well as one
ordered logistic regression model will be considered. Subsequently, the
bias in the parameters and the \(95\%\) confidence interval coverage of
the synthetic data will be examined. Similar to multiple imputation of
missing data, correct inferences from synthetic data require correct
pooling over the multiply imputed data sets.

@@

Also similarly to the missing data case, variances, and subsequently
confidence intervals, should incorporate the increase in variance that
is due to imputation
\citep{reiter_partially_inference_2003, drechsler_synthetic_2011}. Yet,
the increase in variance due to imputation differs according to whether
missing values are imputed or observed data is overimputed with missing
values. Whereas the variance estimate after imputation of missing data
needs to account for the fact that a certain amount of information in
the data is missing, variance estimation from synthetic data does not
suffer from this issue. The adjusted variance estimate that follows from
using multiple synthetic data sets only suffers from the fact that a
finite number of \(m\) synthetic data sets are used to resemble the
observed data. Hence, the according variance estimate for synthetic data
as developed by \citet{reiter_partially_inference_2003} yields \[
T = \bar{u}_m + \frac{b_m}{m},
\]

@@

\begin{table}[ht]
\caption{Univariate descriptives for the true data and $m=5$ pooled univariate descriptives for the synthetic data over 1000 simulations. Variable names followed by a $^*$ are categorical.}
\centering
\begin{tabular}{rrrrrrrrr}
  \hline
 & n & mean & sd & median & min & max & skew & kurtosis \\ 
  \hline
original age & 748 & 9.16 & 6.89 & 10.50 & 0.04 & 21.18 & -0.03 & -1.56 \\ 
  synthetic age & 748 & 9.15 & 6.89 & 10.49 & 0.04 & 20.96 & -0.03 & -1.55 \\ 
  original hgt & 748 & 131.10 & 46.52 & 145.75 & 50.00 & 198.00 & -0.30 & -1.47 \\ 
  synthetic hgt & 748 & 131.06 & 46.50 & 145.32 & 50.69 & 197.16 & -0.30 & -1.47 \\ 
  original wgt & 748 & 37.12 & 26.03 & 34.55 & 3.14 & 117.40 & 0.38 & -1.03 \\ 
  synthetic wgt & 748 & 37.09 & 26.00 & 34.44 & 3.35 & 112.26 & 0.38 & -1.03 \\ 
  original bmi & 748 & 18.04 & 3.04 & 17.45 & 11.73 & 31.74 & 1.14 & 1.79 \\ 
  synthetic bmi & 748 & 18.05 & 3.08 & 17.48 & 11.49 & 32.37 & 1.11 & 1.85 \\ 
  original hc & 748 & 51.62 & 5.86 & 53.10 & 33.70 & 65.00 & -0.91 & 0.12 \\ 
  synthetic hc & 748 & 51.61 & 5.86 & 53.18 & 34.38 & 62.85 & -0.91 & 0.12 \\ 
  original gen* & 748 & 2.53 & 1.59 & 2.00 & 1.00 & 5.00 & 0.52 & -1.36 \\ 
  synthetic gen* & 748 & 2.53 & 1.59 & 2.00 & 1.00 & 5.00 & 0.52 & -1.35 \\ 
  original phb* & 748 & 2.75 & 1.86 & 2.00 & 1.00 & 6.00 & 0.56 & -1.25 \\ 
  synthetic phb* & 748 & 2.75 & 1.86 & 2.00 & 1.00 & 6.00 & 0.56 & -1.24 \\ 
  original tv & 748 & 8.43 & 8.12 & 3.00 & 1.00 & 25.00 & 0.85 & -0.78 \\ 
  synthetic tv & 748 & 8.42 & 8.11 & 3.19 & 1.00 & 25.00 & 0.85 & -0.77 \\ 
  original reg* & 748 & 3.02 & 1.14 & 3.00 & 1.00 & 5.00 & -0.08 & -0.77 \\ 
  synthetic reg* & 748 & 3.02 & 1.14 & 3.00 & 1.00 & 5.00 & -0.08 & -0.76 \\ 
   \hline
\end{tabular}
\end{table}

We see from Table 1 that the synthetic data estimates closely resemble
the true data estimates. All sample statistics of interest show
negligible bias over the 1000 synthetic data sets. Hence, univariately
the imputation model proves adequate.

@@

We see that the correlations obtained from synthetic data are unbiased
with respect to the true data set. The largest absolute bias over 1000
simulations equals 0.013, indicating that the imputation model is
capable of preserving the bivariate relations in the data.

@@

\begin{table}[ht]
\caption{Simulation results for a linear regression model with continuous and ordered categorical predictors. The model evaluated is \texttt{hgt $\sim$ age + phb}. Depicted are the true data estimate and the bias from the true data estimate and the coverage rate of the 95\% confidence interval for the bootstrap and synthetic data sets.}
\centering
\begin{tabular}{rrrrrr}
  \hline
  & & \multicolumn{2}{c}{Bootstrap} & \multicolumn{2}{c}{Synthetic}\\
 term & estimate & bias & cov & bias & cov \\ 
  \hline
(Intercept) & 63.087 & -0.001 & 0.970 & 0.405 & 0.958 \\ 
  age & 7.174 & 0.000 & 0.958 & -0.033 & 0.947 \\ 
  phb.L & -12.250 & 0.008 & 0.950 & 0.582 & 0.927 \\ 
  phb.Q & -1.376 & -0.022 & 0.926 & 0.112 & 0.934 \\ 
  phb.C & -3.564 & 0.051 & 0.915 & 0.301 & 0.912 \\ 
  phb\verb|^|4 & -0.431 & 0.016 & 0.930 & 0.106 & 0.940 \\ 
  phb\verb|^|5 & 2.064 & 0.060 & 0.941 & 0.077 & 0.943 \\ 
   \hline
\end{tabular}
\end{table}

We see that the finite nature of the true data set together with the
design-based simulation setup yields slight undercoverage for the dummy
variables of \texttt{phb}. This finding is observed in both the
bootstrap coverages (i.e.~the fraction of 95\% confidence intervals that
cover the true data parameters) and the synthetic data coverages. Hence,
it is likely that this undercoverage stems from the simulation setup,
rather than the imputation procedure. Besides the undercoverage, there
is a tiny bit of bias in the estimated coefficients of the variable
\texttt{phb} that occurs in the synthetic estimates, but not in the
observed estimates. Yet, since the bias is relatively small and does not
result in confidence invalidity, it seems fair to assume that the
introduced bias is not that problematic.

@@

\begin{table}[H]
\caption{Simulation results for a proportional odds logistic regression model with continuous and ordered categorical predictors. The model evaluated is \texttt{gen $\sim$ age + hc + reg}. Depicted are the true data estimate and the bias from the true data estimate and the coverage rate of the 95\% confidence interval for the bootstrap and synthetic data sets.}
\centering
\begin{tabular}{rrrrrr}
  \hline
 & & \multicolumn{2}{c}{Bootstrap} & \multicolumn{2}{c}{Synthetic}\\
 term & estimate & bias & cov & bias & cov \\ 
  \hline
  age & 0.461 & 0.004 & 0.942 & 0.002 & 0.939 \\
  hc & -0.188 & -0.000 & 0.929 & -0.004 & 0.945 \\
  regeast & -0.339 & 0.012 & 0.960 & 0.092 & 0.957\\
  regwest & 0.486 & 0.009 & 0.952 & -0.122 & 0.944\\
  regsouth & 0.646 & 0.012 & 0.966 & -0.152 & 0.943 \\
  regcity & -0.069 & 0.012 & 0.940 & 0.001 & 0.972 \\
  G1$|$G2 & -6.322 & 0.032 & 0.934 & -0.254 & 0.946 \\
  G2$|$G3 & -4.501 & 0.052 & 0.936 & -0.246 & 0.945 \\
  G3$|$G4 & -3.842 & 0.058 & 0.937 & -0.244 & 0.948 \\
  G4$|$G5 & -2.639 & 0.064 & 0.936 & -0.253 & 0.947 \\
   \hline
\end{tabular}
\end{table}

These results demonstrate that the synthetic data analysis yields
inferences that are on par with inferences from the analyses directly on
the bootstrapped datasets. Hence, for the regression coefficients as
well as the intercepts, the analyses on the synthetic data yield valid
results. Nevertheless, a small amount of bias is introduced in the
estimated intercepts of the synthetic data. However, the corresponding
confidence interval coverage rates are actually somewhat higher than the
confidence interval coverage rates of the bootstrapped data. Therefore,
the corresponding inferences do not seem to be affected by this small
bias.

@@

We demonstrate that generating synthetic data sets with \texttt{mice} in
\texttt{R} is a straightforward process that fits well in a data
analysis pipeline. The approach is hardly different from using multiple
imputation to solve problems related to missing data, and hence can be
expected to be familiar to applied researchers. The multiple synthetic
sets yield valid inferences on the true underlying data generating
mechanism, thereby capturing the nature of the original data. This makes
the multiple synthetisation procedure with \texttt{mice} suitable for
further dissemination of synthetic data sets.
